Individual and Differential Harm in Redistricting

### 2022-04-18 paper/paper.Rmd
revise s2

To illustrate the concepts and measures we introduce, we will make use of three examples, which are stylized representations of U.S. states: Mininois, Minissouri, and Minichussets.
Mininois has one large Democratic city which can electorally dominate the rest of the state; Minissouri has two medium-sized cities which are not large enough to deliver a statewide Democratic majority; and Minichusetts has a uniform 3--2 Democratic split in every precinct.
Each example state has the same population and should be divided into 3 districts with population between 54 and 66.
Voters' preferences are assumed to be stable across elections, and turnout is 100%.
Schematics for these examples are shown in Figure \ref{fig:mini-schematic}(a)--(c).

@@

First, we consider the case of Minichusetts.
Modeled on Massachusetts, Minichusetts contains a remarkable lack of partisan segregation.
This high level of integration, combined with Democrats receiving the majority of votes, leads to a Republican inability to win any seats under recent historical voting patterns [@duch:18].
For existing measures, this creates a conflict: Republicans win about 40% of the vote, but zero of the nine seats.
The Efficiency Gap identifies that the enacted map favors Democrats, whereas both Partisan Bias and Mean-Median indicate a pro-Republican bias under reconstituted recent elections.^[This is according to [PlanScore](https://planscore.campaignlegal.org/massachusetts/#!2020-plan-ushouse-eg), run by the Campaign Legal Center.]

Yet, these scores lack a counterfactual.
Functionally, neither Republicans nor Democrats could do better or worse under this plan than any other plan---a counterfactual plan.
By relying on counterfactual plans, we can then identify the cases where individuals are not represented by their preferred candidate or party, as an alternative to existing partisan bias mesaures.
In places like Minichusetts, as no plan can be drawn which improves the partisan representation of Republican voters, under a counterfactual-driven approach, then there is no harm or bias.

@@

Not all cases are as simple as those where one party wins all the seats.
Mininois also has a Democratic majority, but the voters are no longer integrated.
In a familiar pattern of partisan sorting, Democrats are sorted asymmetrically into a single population-dominating city whereas Republicans dominate the less ideologically segregated rural regions [@chen:rodd:13; @enos:brown:21].
This allows for more variation in the expected partisanship of the districts drawn.

@@

This pattern allows us to build out the intuition behind our proposed measure.
Each voter belongs to a group, mostly simply a party.
For this, we offer that a voter is harmed if they are not represented by their party of choice, but have a counterfactual where they would be represented by their party of choice.

Consider the example in Figure \ref{fig:harm-mininois}.
A Democratic gerrymander in Mininois effectuates Democratic control by splitting its city into with the blades of a pinwheel, reaching out into the more Republican and less population-dense remainder of the state.
We see that the Democratic gerrymander would likely elect zero Republicans and three Democrats, whereas the proportional map would likely elect two Democrats and one Republican.
Figure \ref{fig:harm-mininois}, Panel C then illustrates th total harm by district, comparing the gerrymander to the proportional plan. 
Republican voters who are forced into a Democratic district under the gerrymander, but would not be under a proportional plan are deemed harmed.
The Republicans who stay in Democratic districts in the proportional plan are not harmed.

@@

We can also consider a less drastic sorting, where medium cities favor Democrats, but no longer dominate the state politics, such as in Minissouri.
This allows for additional variability, where either party can gerrymander to draw more seats for their party than a neutral process might allow.

This framework easily extends to additional parties or races. For example, we can index for the interaction between White and Black with Democrats and Republicans.
This framework allows for us to index by these groups and simply break down the interactions as components of the sum of the harms.

@@

Figure \ref{fig:mini-harm} illustrates this measure for a set of hypothetical gerrymanders, comparing to the set of optimal plans from Figure .
As might be intuitively expected, total harm to voters is concentrated in areas whose party representation flips from the gerrymander to the counterfactual plan.

In the districting plan examples of Figure \ref{fig:mini-harm}, the differential harm of the Democratic gerrymander of Mininois is $-0.519$, the differential harm of the Democratic gerrymander of Minissouri is $-0.415$, and the differential harm of the Republican gerrymander of Minissouri is $0.450$, where positive values indicate Republican bias.
In other words, the average Republican voter in Mininois is 51.9% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the utility-maximizing plan in Figure.

@@

This section will formalize and generalize the concepts in Section \ref{sec:motivation}, beginning with some notation.

## Setup

Let $\Voters$ be the set of individuals, or voters, in the region and time period under consideration, who vote to select a candidate to represent each of a set of legislative districts, indexed by $\Distrs$.

@@

To that end, let $\Groups=\{G_1,G_2,\dots\}$ be a set of mutually exclusive groups of voters, so that for each voter $i$, we have $i\in G_k$ for some group $G_k$.
In general we will not require the group identities to be fixed with respect to elections.[^1]
This is because some group identities, like partisan preference, may themselves change election-to-election.

[^1]: Formally, we have random indicator variables $\ind\{i\in G_k\}$ for all $i$ and $k$, subject to the constraint that $\sum_k \ind\{i\in G_k\}=1$ for all $i$.
    These indicator variables may be correlated with $V$ and/or $\tilde Q$, in general.

@@

It is this measure of differential harm that we propose as a unified framework for evaluating a districting plan across a variety of dimensions.
It is derived from an individual-level measure of representation and it accounts for the space of counterfactual possibilities.
Importantly it separates *vote choice* from *group membership*, allowing one to discuss how partisan election outcomes under a specific districting plan create harm for specific groups or their interactions, simply by conditioning $H(q)$ on a set of individuals.

@@

We sample 10,000 districting plans using the SMC algorithm of @mccartan2020, setting constraints to favor compact plans with district boundaries that tend to follow county lines.
One such plan is shown in Figure \ref{fig:nj-maps}(d).
For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm versus the simulated baseline,[^2] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
[^3] The distribution of these measures, along with their pairwise relationships, are shown in Figure \ref{fig:nj-pairs}.

[^2]: As in section \ref{sec:connections}, this is $\Delta_{\dem-\rep}H(q)$, so that positive values correspond to a plan that favors Republicans.

[^3]: We include partisan bias and mean-median in these figures as it is common practice, though it may have theoretical issues when used for a delegation rather than a legislature [@katz21].

@@

In practice, districting plans are built from discrete geographic units (often precincts), and election returns are reported at this level as well.[^4]
Throughout this appendix, let $j$ index the set of electoral precincts or other units.
For concreteness here and without loss of generality we will use the Democratic party as the reference party, letting $s_{jv}$ denote the Democratic two-party vote share in precinct $j$ in election $v$.
Write-in or minor third-party candidates can be safely ignored, since the overall winner will be precisely the two-party vote winner.

[^4]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
    Population information comes from Census blocks, which are practically never split.

@@

The proportionate uniform shift model assumes that from one election to another, a constant fraction of *voters of one party* in each precinct switch to the other party, and that per-precinct turnout is constant across elections.[^5]
This has several advantages over the uniform swing model, where a constant fraction of *all voters* shifts parties.
First, the proportionate shift prevents hypothetical party vote shares from exceeding 1 or falling below 0 in any precinct.
Second, as it has the effect of scaling vote shares across precincts by a constant multiple, it preserves the relative balance of a party's voters across precincts.
For $\Delta_{\dem-\rep}H(q)$, which considers voters in each party separately, this helpfully simplifies computations.

[^5]: Everything carries through if one allows for shifts in the overall turnout level across elections, as long as (a) the relative turnout across precincts is fixed, and (b) turnout shifts are independent of shifts in party preferences.

The proportionate shift may be mathematically expressed as linear shift on a transformed scale, not unlike a generalized linear model.
Specifically, we define $\gamma:[0,1]\to\R$, $$
\gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
$$ which is the inverse CDF of a Laplace distribution with rate 2.[^6]
Then the Democratic vote share in precinct $P_j$ and election $v$, notated here as $s_{jv}$, is modeled as $$
\gamma(s_{jv}) = \gamma(\overline s_j) + \eps_v, \qquad \eps_v \iid \Norm(0, \sigma^2_{\mathrm{hist}}),
$$ where $\overline{s}_j$ is the party baseline in the precinct, and $\sigma^2_\mathrm{hist}$ is estimated from the historical variability in nationwide popular vote shares.
For congressional elections in the United States from 1974--2018, we have $\widehat\sigma^2_\mathrm{hist}=0.03723$, and so draw predictive deviations from $\mathrm{t}_{22}(0, \widehat\sigma^2_\mathrm{hist})$.

[^6]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.

It will simplify matters following to have notation for the overall Democratic vote share in a district.
For a precinct $j$, let $q^{-1}(j)$ denote the district it belongs to. Then let \[
    S_{q^{-1}(j), v}=\frac{\sum_{l\in q^{-1}(j)} s_{lv}\avgvote_l}{\sum_{l\in q^{-1}(j)} \avgvote_l}
\] denote the vote share in the district, where $\avgvote_l$ counts the total number of votes in precinct $l$ at baseline.
The baseline vote share in the district is then $\bar S_{q^{-1}(j)}$.

@@

\begin{equation}\label{eq:appharm1}
\Pr(\win_{q(I)}(V, q)=\rep\text{ and }\win_{\tilde q(I)}(V, \tilde q)=\dem\mid I\in j) 
= \Pr\qty(S_{q^{-1}(j),V} < \half, S_{\tilde q^{-1}(j),V} > \half).
\end{equation}

Because of the uniform shift, for any election $v$, $S_{q^{-1}(j),v} - S_{\tilde q^{-1}(j),v}$ is constant.
Therefore if $S_{q^{-1}(j),v}>S_{\tilde q^{-1}(j),v}$, this probability is zero (if a Democrat wins under $\tilde q$ they will definitely win under $q$, which has a higher Democratic vote share).
But if $S_{q^{-1}(j),v}<S_{\tilde q^{-1}(j),v}$, we can rewrite \eqref{eq:appharm1} as the probability that $\eps_{V}$ falls in a particular range.
Specifically, we have in general that

\begin{equation}\label{eq:appharm2}
\Pr(\win_{q(I)}(V, q)=\rep\text{ and }\win_{\tilde q(I)}(V, \tilde q)=\dem\mid I\in j) 
= \qty(\Phi\qty(\frac{\gamma(\bar S_{\tilde q^{-1}(j)})}{\sigma^2_\mathrm{hist}})
- \Phi\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}))_+,
\end{equation}

where $(\cdot)_+\dfeq \max\{0, \cdot\}$.
If we condition on $V(I)=\dem$, then \eqref{eq:appharm2} is exactly the expected harm to voter $I$ under plans $q$ and $\tilde q$.
The calculation is analogous, up to sign changes, for the expected harm conditioning on $V(I)=\rep$.
This completes the first step: averaging harm within precinct and groups and across elections.

We can then estimate the average across counterfactual plans $\tilde Q$ with the usual Monte Carlo estimator: \[
    \hat h_{j,\dem}(q) 
    \dfeq \frac{1}{N_\text{sim}} \sum_{k=1}^{N_\text{sim}} 
    \qty(\Phi\qty(\frac{\gamma(\bar S_{\tilde q_k^{-1}(j)})}{\sigma^2_\mathrm{hist}})
        - \Phi\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}))_+,
\] where $\hat h_{j,\dem}(q)$ is our estimate of the average harm to a Democratic voter in precinct $j$, averaging across $V$ and $\tilde Q$. The corresponding $\hat h_{j,\rep}(q)$ is computed similarly.

@@

\begin{align*}
\hat H(q) &\dfeq
\frac{\sum_{j} (\bar s_j\hat h_{j,\dem}(q) + 
(1-\bar s_j)\hat h_{j,\rep}(q))\avgvote_j}{\sum_{j} \avgvote_j} \qand \\
\Delta_{\dem-\rep}\hat H(q) &\dfeq
\frac{\sum_{j} (\hat h_{j,\dem}(q) - \hat h_{j,\rep}(q))\avgvote_j}{\sum_{j} \avgvote_j}.
\end{align*}

Here, we rely on the fact that even across election shifts, the relative balance of each party's voters across precincts is preserved.

>>>>

To illustrate the concepts and measures we introduce, we will make use of three examples, which are stylized representations of U.S. states: Mininois, Minissouri, and Minichusetts.
Mininois has one large Democratic city which can electorally dominate the rest of the state; Minissouri has two medium-sized cities which are not large enough to deliver a statewide Democratic majority; and Minichusetts has a uniform 3--2 Democratic split in every precinct.
Each example state has the same population and should be divided into 3 districts with population between 54 and 66.
Voters' preferences are assumed to be stable across elections, and turnout is 100%.
Schematics for these examples are shown in Figure \ref{fig:mini-schematic}(a)--(c).

@@

Since every precinct in Minichusetts has a Democratic majority, it is impossible to draw a districting plan built from precincts that creates even a single Republican district.
This mirrors the situation in the actual state of Massachusetts, where a high level of partisan integration and a solid statewide Democratic majority leads to Republican inability to win any congressional seats under recent historical voting patterns [@duch:18].

This Democratic dominance poses a problem for existing measures of partisan gerrymandering.
One of the most common measures, the efficiency gap of @stephanopoulos2015, takes a value of $-0.311$ or $-0.317$ for Minichusetts, depending on the configuration of the districts.
This would make any Minichusetts districting plan an extreme Democratic gerrymander in comparison to the historical range of efficiency gaps observed in states nationwide.[^1]

[^1]: See, e.g., the historical distribution calculated by the Campaign Legal Center's PlanScore tool: [planscore.campaignlegal.org/metrics/efficiencygap/](https://planscore.campaignlegal.org/metrics/efficiencygap/).

But it is impossible to gerrymander if every manipulation of district boundaries produces no change in partisan outcomes.
What the efficiency gap and other gerrymandering measures lack is an awareness of the space of possible districting plans, plans that *counterfactually* might have been selected.
Once a set of districting rules has been agreed on (in Minichusetts, that districts must be contiguous and composed of whole precincts), what matters is not how a plan measures up to a fixed ideal, but how it performs in comparison to counterfactual plans.
There can be no gerrymander, and no harm to individual voters, if no improvement in partisan fairness is possible.

@@

To explore how to properly integrate counterfactual districting plans into the analysis of a potential gerrymander, we turn to the more interesting state of Mininois, where, in a familiar pattern of partisan sorting, Democrats are sorted asymmetrically into a single population-dominating city whereas Republicans dominate the less ideologically segregated rural regions [@chen:rodd:13; @enos:brown:21].

Consider the two plans in Figure \ref{fig:harm-mininois}(a)-(b).
A Democratic gerrymander (a) effectuates Democratic control by splitting its city into with the blades of a pinwheel, reaching out into the more Republican and less population-dense remainder of the state.
We see that the Democratic gerrymander would elect zero Republicans and three Democrats, whereas the proportional map (b) would elect two Democrats and one Republican.

@@

Suppose now that plan (a) is enacted into law.
Who does this decision harm, if anyone?
We offer that a voter is harmed if they are not represented by their party of choice, but have a counterfactual where they would be represented by their party of choice.
Using plan (b) as a counterfactual, we can see that rural Republicans in the eastern half of Mininois are harmed by plan (a), since under plan (b) they would be represented by a Republican.
Republicans in the city and in the southwest are unharmed, as are all Democratic voters,, since they are represented by the same party under both plans.
Figure \ref{fig:harm-mininois}, Panel C illustrates this by plotting the number of voters harmed in each precinct.

Measuring the harm of the Democratic gerrymander required comparing it to a fair counterfactual plan.
The choice of counterfactual might seem hopelessly subjective.
We discuss in Section \ref{sec:estimating} principled ways to define counterfactuals, but for now we hope the value in measuring harm through counterfactuals is clear even if the operationalization is not.

@@

We could quantify the harm a Democratic gerrymander imposed on voters in Mininois by comparing election outcomes to those under a counterfactual proportional plan.
But our results said nothing about the direction of the harm.
While it is apparent that the Democratic gerrymander indeed benefits Democrats, tallying up the total harm done to all voters is not enough to prove this.

However, we can partition voters into groups, and compare how the harm done by a plan falls across those groups.
If a plan disproportionately harms voters of one group, then we might conclude that that group was the target of a gerrymander.
Figure \ref{fig:diffharm-minissouri} illustrates this idea for two hypothetical gerrymanders ((a) and (c)) of Minissouri.
As in Mininois, we can tally the total number of voters harmed in each precinct ((d) and (f)) in comparison to a proportional counterfactual plan (b).
As might be intuitively expected, total harm to voters is concentrated in areas whose party representation flips from the gerrymander to the counterfactual plan.

@@

Panel (e) then splits the voters by party, and computes, for each party, the share of its voters which harmed.
The Democratic gerrymander (a) harms Republicans at a much higher rate than Democrats, and the opposite can be said of the Republican gerrymander (c).
We call the gap between these two rates the *differential harm,* and propose it as a general way to account for the impact of a districting plan on particular groups.

The differential harm of the Democratic gerrymander of Minissouri is $-0.415$, and the differential harm of the Republican gerrymander is $0.450$, where positive values indicate Republican bias.
In other words, under the Democratic gerrymander, the average Republican voter in Mininois is 41.5% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the proportional plan in Figure \ref{fig:diffharm-minissouri}(b).

@@

This section will formalize and generalize the concepts in Section \ref{sec:motivation}, beginning with some notation.
Let $\Voters$ be the set of individuals, or voters, in the region and time period under consideration, who vote to select a candidate to represent each of a set of legislative districts, indexed by $\Distrs$.

@@

To that end, let $\Groups=\{G_1,G_2,\dots\}$ be a set of mutually exclusive groups of voters, so that for each voter $i$, we have $i\in G_k$ for some group $G_k$.
In general we will not require the group identities to be fixed with respect to elections.[^2]
This is because some group identities, like partisan preference, may themselves change election-to-election.

[^2]: Formally, we have random indicator variables $\ind\{i\in G_k\}$ for all $i$ and $k$, subject to the constraint that $\sum_k \ind\{i\in G_k\}=1$ for all $i$.
    These indicator variables may be correlated with $V$ and/or $\tilde Q$, in general.

@@

It is this measure of differential harm that we propose as a unified framework for evaluating a districting plan across a variety of dimensions.
It is derived from an individual-level measure of representation and it accounts for the space of counterfactual possibilities.
Importantly, it separates *vote choice* from *group membership*, allowing one to discuss how partisan election outcomes under a specific districting plan create harm for specific groups or their interactions, simply by conditioning $H(q)$ on a set of individuals.

@@

We sample 10,000 districting plans using the SMC algorithm of @mccartan2020, setting constraints to favor compact plans with district boundaries that tend to follow county lines.
One such plan is shown in Figure \ref{fig:nj-maps}(d).
For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm versus the simulated baseline,[^3] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
[^4] The distribution of these measures, along with their pairwise relationships, are shown in Figure \ref{fig:nj-pairs}.

[^3]: As in section \ref{sec:connections}, this is $\Delta_{\dem-\rep}H(q)$, so that positive values correspond to a plan that favors Republicans.

[^4]: We include partisan bias and mean-median in these figures as it is common practice, though it may have theoretical issues when used for a delegation rather than a legislature [@katz21].

@@

In practice, districting plans are built from discrete geographic units (often precincts), and election returns are reported at this level as well.[^5]
Throughout this appendix, let $j$ index the set of electoral precincts or other units.
For concreteness here and without loss of generality we will use the Democratic party as the reference party, letting $s_{jv}$ denote the Democratic two-party vote share in precinct $j$ in election $v$.
Write-in or minor third-party candidates can be safely ignored, since the overall winner will be precisely the two-party vote winner.

[^5]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
    Population information comes from Census blocks, which are practically never split.

@@

The proportionate uniform shift model assumes that from one election to another, a constant fraction of *voters of one party* in each precinct switch to the other party, and that per-precinct turnout is constant across elections.[^6]
This has several advantages over the uniform swing model, where a constant fraction of *all voters* shifts parties.
First, the proportionate shift prevents hypothetical party vote shares from exceeding 1 or falling below 0 in any precinct.
Second, as it has the effect of scaling vote shares across precincts by a constant multiple, it preserves the relative balance of a party's voters across precincts.
For $\Delta_{\dem-\rep}H(q)$, which considers voters in each party separately, this helpfully simplifies computations.

[^6]: Everything carries through if one allows for shifts in the overall turnout level across elections, as long as (a) the relative turnout across precincts is fixed, and (b) turnout shifts are independent of shifts in party preferences.

The proportionate shift may be mathematically expressed as linear shift on a transformed scale, not unlike a generalized linear model.
Specifically, we define $\gamma:[0,1]\to\R$, $$
\gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
$$ which is the inverse CDF of a Laplace distribution with rate 2.[^7]
Then the Democratic vote share in precinct $P_j$ and election $v$, notated here as $s_{jv}$, is modeled as $$
\gamma(s_{jv}) = \gamma(\overline s_j) + \eps_v, \qquad \eps_v \iid \Norm(0, \sigma^2_{\mathrm{hist}}),
$$ where $\overline{s}_j$ is the party baseline in the precinct, and $\sigma^2_\mathrm{hist}$ is estimated from the historical variability in nationwide popular vote shares.
For congressional elections in the United States from 1974--2018, we have $\widehat\sigma^2_\mathrm{hist}=0.03723$, and so draw predictive deviations from $\mathrm{t}_{22}(0, \widehat\sigma^2_\mathrm{hist})$.

[^7]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.

It will simplify matters following to have notation for the overall Democratic vote share in a district.
For a precinct $j$, let $q^{-1}(j)$ denote the district it belongs to.
Then let $$
    S_{q^{-1}(j), v}=\frac{\sum_{l\in q^{-1}(j)} s_{lv}\avgvote_l}{\sum_{l\in q^{-1}(j)} \avgvote_l}
$$ denote the vote share in the district, where $\avgvote_l$ counts the total number of votes in precinct $l$ at baseline.
The baseline vote share in the district is then $\bar S_{q^{-1}(j)}$.

@@

```{=tex}
\begin{equation}\label{eq:appharm1}
\Pr(\win_{q(I)}(V, q)=\rep\text{ and }\win_{\tilde q(I)}(V, \tilde q)=\dem\mid I\in j) 
= \Pr\qty(S_{q^{-1}(j),V} < \half, S_{\tilde q^{-1}(j),V} > \half).
\end{equation}
```
Because of the uniform shift, for any election $v$, $S_{q^{-1}(j),v} - S_{\tilde q^{-1}(j),v}$ is constant.
Therefore if $S_{q^{-1}(j),v}>S_{\tilde q^{-1}(j),v}$, this probability is zero (if a Democrat wins under $\tilde q$ they will definitely win under $q$, which has a higher Democratic vote share).
But if $S_{q^{-1}(j),v}<S_{\tilde q^{-1}(j),v}$, we can rewrite \eqref{eq:appharm1} as the probability that $\eps_{V}$ falls in a particular range.
Specifically, we have in general that

```{=tex}
\begin{equation}\label{eq:appharm2}
\Pr(\win_{q(I)}(V, q)=\rep\text{ and }\win_{\tilde q(I)}(V, \tilde q)=\dem\mid I\in j) 
= \qty(\Phi\qty(\frac{\gamma(\bar S_{\tilde q^{-1}(j)})}{\sigma^2_\mathrm{hist}})
- \Phi\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}))_+,
\end{equation}
```
where $(\cdot)_+\dfeq \max\{0, \cdot\}$.
If we condition on $V(I)=\dem$, then \eqref{eq:appharm2} is exactly the expected harm to voter $I$ under plans $q$ and $\tilde q$.
The calculation is analogous, up to sign changes, for the expected harm conditioning on $V(I)=\rep$.
This completes the first step: averaging harm within precinct and groups and across elections.

We can then estimate the average across counterfactual plans $\tilde Q$ with the usual Monte Carlo estimator: $$
    \hat h_{j,\dem}(q) 
    \dfeq \frac{1}{N_\text{sim}} \sum_{k=1}^{N_\text{sim}} 
    \qty(\Phi\qty(\frac{\gamma(\bar S_{\tilde q_k^{-1}(j)})}{\sigma^2_\mathrm{hist}})
        - \Phi\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}))_+,
$$ where $\hat h_{j,\dem}(q)$ is our estimate of the average harm to a Democratic voter in precinct $j$, averaging across $V$ and $\tilde Q$.
The corresponding $\hat h_{j,\rep}(q)$ is computed similarly.

@@

```{=tex}
\begin{align*}
\hat H(q) &\dfeq
\frac{\sum_{j} (\bar s_j\hat h_{j,\dem}(q) + 
(1-\bar s_j)\hat h_{j,\rep}(q))\avgvote_j}{\sum_{j} \avgvote_j} \qand \\
\Delta_{\dem-\rep}\hat H(q) &\dfeq
\frac{\sum_{j} (\hat h_{j,\dem}(q) - \hat h_{j,\rep}(q))\avgvote_j}{\sum_{j} \avgvote_j}.
\end{align*}
```
Here, we rely on the fact that even across election shifts, the relative balance of each party's voters across precincts is preserved.
